---
title: "Impact of Lifestyle and Physical Indicators on First Diagnosis of Cancer"
author:
  - Cong Lyu
  - Christine Xing
  - Shahzab Hussain
output:
  pdf_document: default
  html_document:
    df_print: paged
---



## Import data

```{r}
library(ipumsr)

ddi <- read_ipums_ddi("nhis_00003.xml")
data <- read_ipums_micro(ddi)
```


```{r}
names(data)
col.ages <- c("CNBRANAG", "CNCOLNAG", "CNCOLRECTAG", "CNESOPAG", "CNHDNCKAG",
              "CNLIVRAG", "CNPANCAG", "CNSTOMAG")
col.cancers <- c("CNBRAN", "CNCOLN", "CNCOLRECT", "CNESOP", "CNHDNCK", 
                 "CNLIVR", "CNPANC", "CNSTOM")
col.numeric.categorical <- c("EDUC", "MAXEDUC", "SPOUSEDUC")
col.categorical <- c("SEX", "QTCINCFAM")
col.key <- c("SERIAL", "NHISHID", "NHISPID", "HHX")
```

## Topic 1 

Response: The age of having cancer for the first time among these kinds of cancer

### EDA 

```{r}
library(dplyr)
# Data cleaning: figure out missing values
data.clean <- data %>%
  mutate(across(col.ages, ~if_else(. >= 96, NA_real_, .))) %>%
  # mutate(across(c("EMPSTAT"), ~if_else(. >= 900, NA_real_, .))) %>%
  # mutate(across(c("HOURSWRK"), ~if_else(. >= 97, NA_real_, .))) %>%
  # mutate(across(c("EMPHI", "EMPFT", "SPOUSWKFT", "HEALTH", "ALCDRINKEV"),
  #               ~if_else(. >= 7, NA_real_, .))) %>%
  mutate(across(c("DIABTYPE"), ~if_else(. >= 7, NA_real_, .))) %>%
  mutate(across(c("HEIGHT"), ~if_else(. >= 95, NA_real_, .))) %>%
  mutate(across(c("WEIGHT", "BMICALC"), ~if_else(. >= 995, NA_real_, .))) %>%
  mutate(across(c("ALCDRINKEV", "ALCEV30D"), ~if_else(. >= 7, NA_real_, .))) %>%
  mutate(across(c("ALCEV30D"), ~if_else(. ==0, 1, .))) %>%
  mutate(across(c("ALC5UPOCC30D"), ~if_else(. >= 97, NA_real_, .))) %>%
  mutate(across(c("ALC5UPEVYR"),  ~if_else((. >= 7) | (. ==0), NA_real_, .))) %>%
  mutate(across(c("MOD10FNO"), ~if_else(. >= 995, NA_real_, .))) %>%
  mutate(across(c("MOD10FWK"), ~if_else((. >= 94) & (. <=96), 0, .))) %>%
  mutate(across(c("MOD10FWK"), ~if_else((. ==93) | (. >= 97), NA_real_, .))) %>%
  mutate(across(c("EMPSTAT"), ~if_else(. >= 900, NA_real_, .))) %>%
  mutate(across(c("EMPHI"), ~if_else((. >= 7) | (. ==0), NA_real_, .))) %>%
  mutate(across(c("SPOUSEDUC"), ~if_else((. >= 97) | (. ==0), NA_real_, .))) 


### Check missing values 
colSums(is.na(data.clean)) 

subset.age.clean =  data.clean %>% 
  filter(!is.na(CNBRANAG) | !is.na(CNCOLNAG) | !is.na(CNCOLRECTAG) |
           !is.na(CNESOPAG) | !is.na(CNHDNCKAG) | !is.na(CNLIVRAG) |
           !is.na(CNPANCAG) | !is.na(CNSTOMAG)) %>%
  mutate(CNAG = pmin(as.numeric(CNBRANAG), as.numeric(CNCOLNAG), 
                     as.numeric(CNCOLRECTAG), as.numeric(CNESOPAG),
                     as.numeric(CNHDNCKAG), as.numeric(CNLIVRAG),
                     as.numeric(CNPANCAG), as.numeric(CNSTOMAG), 
                     na.rm = TRUE)) %>% 
  mutate(across(c(SEX, EMPSTAT, EMPHI, QTCINCFAM, 
                  ALCDRINKEV, ALC5UPEVYR, ALCEV30D, # Binary variables
                  DIABTYPE), # non-numerical factor
                ~as_factor(as_factor(.))))

```

Check info and labels

```{r}
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(SEX), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(MAXEDUC), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(SPOUSEDUC), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(EMPSTAT), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(EMPHI), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(QTCINCFAM), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(HEALTH), y=CNAG)) + ggplot2::geom_boxplot()

ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = HEIGHT, y=CNAG)) + ggplot2::geom_point()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = WEIGHT, y=CNAG)) + ggplot2::geom_point()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = BMICALC, y=CNAG)) + ggplot2::geom_point()

ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(DIABTYPE), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(ALCDRINKEV), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(ALC5UPEVYR), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(ALCEV30D), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(ALC5UPOCC30D), y=CNAG)) + ggplot2::geom_boxplot()
# ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(MOD10FNO), y=CNAG)) + ggplot2::geom_boxplot()
ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(MOD10FWK), y=CNAG)) + ggplot2::geom_boxplot()

hist(subset.age.clean$CNAG, breaks=43)
head(data$MOD10FWK   )
summary(data$MOD10FWK)

```

### Models

#### Saturated model

```{r}
library(MASS)
subset.age.clean.nb <- subset.age.clean %>%
  dplyr::select(SEX, MAXEDUC, EMPSTAT, QTCINCFAM, HEALTH, HEIGHT, 
                WEIGHT, BMICALC, DIABTYPE, ALCEV30D, MOD10FWK, CNAG) %>%
  na.omit()

fit0.poisson = glm(CNAG ~ SEX + MAXEDUC + EMPSTAT + QTCINCFAM 
           + HEALTH + HEIGHT + WEIGHT + BMICALC + DIABTYPE 
           + ALCEV30D + MOD10FWK,
           data=subset.age.clean.nb, family=poisson(link="log"),
           na.action = na.omit)
anova(fit0.poisson)
summary(fit0.poisson)
```
#### Model selection
Simplify: Backward stepwise

```{r}
fit0.poisson.backward <- step(fit0.poisson, direction = "backward")
summary(fit0.poisson.backward)
anova(fit0.poisson, fit0.poisson.backward)
```

```{r}
anova(fit0.poisson, fit0.poisson.backward, test="Chisq")
```

```{r}
# Load the necessary libraries
library(ggplot2)
library(reshape2)

# Extract coefficients from both models
coefficients_saturated <- coef(fit0.poisson)
coefficients_reduced <- coef(fit0.poisson.backward)

# Remove the intercepts
coefficients_saturated <- coefficients_saturated[-1]
coefficients_reduced <- coefficients_reduced[-1]

# Create data frames for plotting
coeff_df_saturated <- data.frame(
  Variable = names(coefficients_saturated),
  Coefficient = coefficients_saturated,
  Model = 'Saturated'
)

coeff_df_reduced <- data.frame(
  Variable = names(coefficients_reduced),
  Coefficient = coefficients_reduced,
  Model = 'Reduced'
)

# Combine the two data frames, allowing NAs for the variables not in the reduced model
coeff_df_combined <- merge(coeff_df_saturated, coeff_df_reduced, by = "Variable", all = TRUE, suffixes = c(".Saturated", ".Reduced"))

# Create a long format data frame for plotting with ggplot2
coeff_df_long <- melt(coeff_df_combined, id.vars = "Variable", 
                      measure.vars = c("Coefficient.Saturated", "Coefficient.Reduced"),
                      variable.name = "Model", value.name = "Coefficient")

# Sort the data frame by the value of the coefficients from the saturated model
coeff_df_long <- coeff_df_long[order(-abs(coeff_df_long$Coefficient)),]

# Reorder the variables based on the sorted coefficients for plotting
coeff_df_long$Variable <- factor(coeff_df_long$Variable, levels = unique(coeff_df_long$Variable))

# Plot using ggplot2
ggplot(coeff_df_long, aes(x = Variable, y = Coefficient, fill = Model)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  coord_flip() +  # Flipping coordinates to make it easier to read variable names
  scale_fill_manual(values = c("Coefficient.Saturated" = "blue", "Coefficient.Reduced" = "green", "NA" = "grey")) +  # Use blue and green for different models
  labs(x = "Variable", y = "Coefficient Value", fill = "Model") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "bottom")  # Adjust the angle of x-axis labels if needed
```

Findings:



#### Mixed model with clusters regarding of EMPSTAT (employed vs. unemployed)

```{r}
# library(lme4)
# 
# fit0.mixed <- glmer(CNAG ~ SEX + MAXEDUC + SPOUSEDUC + EMPSTAT + EMPHI + 
#   QTCINCFAM + HEALTH + HEIGHT + WEIGHT + BMICALC + DIABTYPE + 
#      ALCEV30D +  MOD10FWK + 
#     (1 | EMPSTAT), 
#   data = subset.age.clean, family = poisson(link = "log"),
#   na.action = na.omit)
# summary(fit0.mixed)
# anova(fit0.mixed)
# 
# fit0.mixed.simplified <- glmer(CNAG ~ SEX+ BMICALC + DIABTYPE + 
#     MOD10FWK   + (1 | EMPSTAT), 
#   data = subset.age.clean, family = poisson(link = "log"),
#   na.action = na.omit)
# summary(fit0.mixed.simplified)
# anova(fit0.mixed, fit0.mixed.simplified)
```
#### Drinking behavior 

```{r}
fit1.AL = glm(CNAG ~ ALCEV30D,
           data=subset.age.clean.nb, family=poisson(link="log"),
           na.action = na.omit)
anova(fit1.AL)
summary(fit1.AL)
```

```{r}
anova(fit0.poisson, fit1.AL, test="Chisq")
```


```{r}
fit1.MOD = glm(CNAG ~ MOD10FWK,
           data=subset.age.clean.nb, family=poisson(link="log"),
           na.action = na.omit)
anova(fit1.MOD)
summary(fit1.MOD)
anova(fit0.poisson, fit1.MOD, test="Chisq")

```

### Negative binomial GLM

The first time to success after Y years is a negative binomial problem.

```{r}
library(dplyr)
library(MASS)

  
colSums(is.na(subset.age.clean.nb))

fit.NB = glm.nb(CNAG ~ SEX + MAXEDUC + EMPSTAT + QTCINCFAM + HEALTH + HEIGHT + 
  WEIGHT + BMICALC + DIABTYPE + ALCEV30D + MOD10FWK,
  data = subset.age.clean.nb)

summary(fit.NB)
anova(fit.NB)

fit.NB.reduced = stepAIC(fit.NB, direction="backward")
summary(fit.NB.reduced)
anova(fit.NB, fit.NB.reduced, test="Chisq")


anova(fit0.poisson, fit.NB, test="Chisq")
anova(fit0.poisson.backward, fit.NB.reduced, test="Chisq")



```


```{r}
# Perform the likelihood ratio test
lrt_result <- anova(fit0.poisson, fit.NB, test = "LRT")

# Print the results of the likelihood ratio test
print(lrt_result)
```






## Topic 2

Response: binary CNLIVR

```{r}
head(data$CNLIVR)
subset.CNLIVR.clean =  data.clean %>% 
  # mutate(across(c("CNLIVR"), ~if_else(. >= 7, NA_real_, .))) %>%
  filter(CNLIVR==1 | CNLIVR==2) %>%
  mutate(CNLIVR=CNLIVR-1) %>% # Change label: Positive 2->1, negative 1->0
  mutate(across(c(SEX, EMPSTAT, EMPHI, QTCINCFAM, 
                  ALCDRINKEV, ALC5UPEVYR, ALCEV30D, # Binary variables
                  DIABTYPE), # non-numerical factor
                ~as_factor(as_factor(.))))
fit.cnliver = glm(CNLIVR ~ AGE + SEX + MAXEDUC + SPOUSEDUC + 
                    EMPSTAT + EMPHI + QTCINCFAM +
                    HEALTH + HEIGHT + WEIGHT + BMICALC + DIABTYPE + 
                    ALCDRINKEV +  ALCEV30D + ALC5UPOCC30D + MOD10FWK,
                  data=subset.CNLIVR.clean, family=binomial(link = "logit"),
                  na.action = na.omit)
anova(fit.cnliver)
summary(fit.cnliver)

```

```{r}
#########################################
## Reference Only, Not for final version
#########################################

#install.packages("glmnet")

library(glmnet)

x <- model.matrix(~ AGE + SEX + MAXEDUC + SPOUSEDUC + 
                    EMPSTAT + EMPHI + QTCINCFAM +
                    HEALTH + HEIGHT + WEIGHT + BMICALC + DIABTYPE + 
                    ALCDRINKEV + ALCEV30D + ALC5UPOCC30D + MOD10FNO - 1, 
                  data=subset.CNLIVR.clean)

yx <- model.matrix(~CNLIVR+ AGE + SEX + MAXEDUC + SPOUSEDUC + 
                     EMPSTAT + EMPHI + QTCINCFAM +
                     HEALTH + HEIGHT + WEIGHT + BMICALC + DIABTYPE + 
                     ALCDRINKEV + ALCEV30D + ALC5UPOCC30D + MOD10FNO - 1, 
                   data=subset.CNLIVR.clean)
y <- yx[,1]

# Lasso regression
fit.cnliver.glmnet <- glmnet(x, y, family="binomial", alpha=1)

# You can then use cv.glmnet to find the optimal lambda value (regularization strength)
fit.optimal <- cv.glmnet(x, y, family="binomial", alpha=1)
# Plot the cross-validation curve
plot(fit.optimal)

```



Correlation check

```{r}
head(data$MOD10FWK)
table(data$MOD10FWK)
# library(corrplot)
# corr_matrix <- cor(subset.age, use=)
# corrplot(corr_matrix, method = "color", addCoef.col = "black", 
#          tl.col="black", tl.srt=45, cl.pos='b', type="upper")

```





```{r}
# 
# hist(subset.age.clean$ALC5UPEVYR)
# ggplot2::ggplot(subset.age.clean, ggplot2::aes(x = as.factor(ALC5UPEVYR), y=CNAG)) + ggplot2::geom_boxplot()


```